## What's a function?

. . .

> A function is a **block of code** organized together to perform a specific task and only runs when it is called. It can have **parameters** and can **return** a result.

<br/>

. . .

{{< fa circle-right >}} &nbsp;**Automate** common and repetitive tasks

. . .

{{< fa circle-right >}} &nbsp;Increase **reproducibility** and **readability** of the code

. . .

<br/><br/>


**Advantages**[^r4ds]

- You can give a function an **evocative name** that makes your code easier to understand.
- As requirements change, you only need to **update code in one place**, instead of many.
- You **eliminate** the chance of making **incidental mistakes** when you copy and paste.
- It makes it **easier to reuse** work from project-to-project, increasing your productivity over time.

[^r4ds]: R for Data Science (2^e^ ed.) <https://r4ds.hadley.nz/functions.html>


## When to write a function? 

<br /><br />

::: {.center}
::: {.boxed}
**Same code reused twice**
<br/>
(do not repeat yourself)
:::


_and/or_


::: {.boxed}
**Reduce code complexity**
<br/>
(workflow modularity)
:::


_and/or_


::: {.boxed}
**Improve code organisation**
<br/>
(workflow readability)
:::

:::



## Writing a function

:::: {.columns}

::: {.column width=45%}
```{r}
#| echo: true
#| eval: false

## Function definition ----

function_name <- function(input) {
  
  # Body 
  # of the function
  
  return(output)
}
```
:::

::: {.column width=10%}
:::

::: {.column width=45%}
- A function is defined by calling **`function()`**
- A function can have a (evocative) **name**[^anonym]
- A function can have 0, 1 or many **arguments** (inputs)
- A function can **return** a value (output) 
:::

::::


[^anonym]: In {{< fa brands r-project >}}&nbsp; there is also **anonymous functions** (also known as **lambda functions**):<br />`function(x) sum(x) / length(x)`<br />or in short: `\(x) sum(x) / length(x)`

. . .


<br/><br/>

{{< fa circle-arrow-right >}}&nbsp; A **two-step** process

<br/>

:::: {.columns}

::: {.column width=45%}
**1. Definition** of the function

```{r}
#| echo: true
#| eval: false

## Function definition ----

arithmetic_mean <- function(x) {
  
  y <- sum(x) / length(x)
  
  return(y)
}
```
:::

::: {.column width=10%}
:::

::: {.column width=45%}
**2. Call** of the function

```{r}
#| echo: true
#| eval: false

## Function call ----

arithmetic_mean(x = c(4, 6, 5, 10))
```
:::

::::



## Exiting a function

{{< fa circle-arrow-right >}}&nbsp; Functions exit in two ways

<br /><br />

:::: {.columns}

::: {.column width=45%}
::: {.center}
[{{< fa check size=2xl >}}]{.green}

<br />

**Success**


i.e. return a value w/ **`return()`**
:::
:::

::: {.column width=10%}
:::

::: {.column width=45%}
::: {.center}
[{{< fa xmark size=2xl >}}]{.red}

<br />

**Failure**


i.e. throw an error w/ **`stop()`**
:::
:::

::::


## Function return

{{< fa circle-arrow-right >}}&nbsp; A function only **returns one single object**

<br />

:::: {.columns}

::: {.column .center width=45%}

```{r}
#| echo: true
#| eval: false

arithmetic_mean <- function(x) {
  
  y <- sum(x) / length(x)
  
  return(y)
}
```
:::

::: {.column .center width=10%}
:::

::: {.column .center width=45%}
:::

::::


. . .

<br />

{{< fa circle-arrow-right >}}&nbsp; Use **`list()`** if you want to return more that one object

<br />

:::: {.columns}

::: {.column .center width=45%}

```{r}
#| echo: true
#| eval: false

function_name <- function(x) {
  
  # ...

  result <- list(
    object_1,
    object_2,
    ...
  )
  
  return(result)
}
```
:::

::: {.column .center width=10%}
:::

::: {.column .center width=45%}
:::

::::




## Function return

{{< fa circle-arrow-right >}}&nbsp; In most cases, you can **omit** the keyword **`return()`**: this is called an **implicit return**

. . .

<br />

:::: {.columns}

::: {.column .center width=33%}
[**Explicit return**]{.small}

```{r}
#| echo: true
#| eval: false

arithmetic_mean <- function(x) {
  
  y <- sum(x) / length(x)
  
  return(y)
}
```
:::

::: {.column .center width=33%}
[**Implicit return**]{.small}

```{r}
#| echo: true
#| eval: false

arithmetic_mean <- function(x) {
  
  y <- sum(x) / length(x)
  
  y
}
```
:::

::: {.column .center width=33%}
:::

::::


## Function return

{{< fa circle-arrow-right >}}&nbsp; In most cases, you can **omit** the keyword **`return()`**: this is called an **implicit return**


<br />

:::: {.columns}

::: {.column .center width=33%}
[**Explicit return**]{.small}

```{r}
#| echo: true
#| eval: false

arithmetic_mean <- function(x) {
  
  y <- sum(x) / length(x)
  
  return(y)
}
```
:::

::: {.column .center width=33%}
[**Implicit return**]{.small}

```{r}
#| echo: true
#| eval: false

arithmetic_mean <- function(x) {
  
  y <- sum(x) / length(x)
  
  y
}
```
:::

::: {.column .center width=33%}
[**Implicit return & simplification**]{.small}

```{r}
#| echo: true
#| eval: false

arithmetic_mean <- function(x) {
  
  sum(x) / length(x)
}
```
:::

::::



## Function return

{{< fa circle-arrow-right >}}&nbsp; But sometimes, you need an **explicit return**

. . .

<br />

:::: {.columns}

::: {.column .center width=45%}
[**Explicit returns**]{.small}

```{r}
#| echo: true
#| eval: false

arithmetic_mean <- function(x) {
  
  if (any(is.na(x))) {

    return(0)
  }

  y <- sum(x) / length(x)
  
  return(y)
}
```
:::

::: {.column .center width=10%}
:::

::: {.column .center width=45%}
[**WRONG behaviour**]{.small}

```{r}
#| echo: true
#| eval: false

arithmetic_mean <- function(x) {
  
  if (any(is.na(x))) {

    0
  }

  sum(x) / length(x)
}
```
:::

::::

. . .

<br />

:::: {.columns}

::: {.column .center width=45%}
:::

::: {.column .center width=10%}
:::

::: {.column .center width=45%}
[**Explicit and implicit returns**]{.small}

```{r}
#| echo: true
#| eval: false

arithmetic_mean <- function(x) {
  
  if (any(is.na(x))) {

    return(0)
  }

  sum(x) / length(x)
}
```
:::

::::


## No return

{{< fa circle-arrow-right >}}&nbsp; Not all functions necessarily return an object

<br />

For instance,

- a function that downloads / saves a file on the disk
- a function that creates a plot (w/ `graphics`)
- a function that modifies a global option
- ...

. . .

<br /><br />

:::: {.columns}

::: {.column width=45%}

{{< fa circle-arrow-right >}}&nbsp; In that case, the function will return **`NULL`**

:::

::: {.column width=10%}
:::

::: {.column width=45%}
```{r}
#| echo: true
#| eval: false

# Define the function ----
save_as_csv <- function(data) {

  write.csv(data, "filename.csv", row.names = FALSE)
}

# Create fake dataset ----
tab <- data.frame(x = 1:10, y = letters[1:10])

# Call the function ----
result <- save_as_csv(data = tab)

result
## NULL
```
:::

::::



## Invisible return

{{< fa circle-arrow-right >}}&nbsp; Sometimes, you might want to make the function's **return invisible** if it is **not assigned** to a variable with the function **`invisible()`**

<br />

This behaviour is _usually_ used: 

- when the function returns **`NULL`**
- when the return is not the main purpose of the function

. . .

<br /><br />


:::: {.columns}

::: {.column width=45%}

{{< fa circle-arrow-right >}}&nbsp; If you don't assigne the output of the function, it will no longer be accessible afterwards

<br />

{{< fa circle-arrow-right >}}&nbsp; If you implement an invisible return, you have to **store the output** of the function in a **variable** and **print this variable**

:::

::: {.column width=10%}
:::

::: {.column width=45%}
```{r}
#| echo: true
#| eval: false

# Define the function ----
save_as_csv <- function(data) {

  file_name <- here::here("outputs", "filename.csv")
  write.csv(data, file_name, row.names = FALSE)

  invisible(file_name)
}

# Call the function (no assignation, no return) ----
save_as_csv(data = tab)

# Call the function (w/ assignation) ----
path <- save_as_csv(data = tab)

path
## [1] '/home/user/documents/project_a/outputs/filename.csv'
```
:::

::::




## To do

- Arguments
- Argument with default value
- Argument order
- ... (dot-dot-dot)
- No argument
- Environments
- Error handling
- Assertive programming
- Calling external packages (`::`)
- Case study: from script to function
- Documenting a function
- Quiz
